From physical principles to ???

### 2016-02-12 2015CompleteDerivation/completeDerivation.tex
Incorporated Dave's comments on first section

% Status: Conceptual work done. Text ready for review.

%TODO: incorporate feedback from Dave. Determinism definition as "relative" to choice of state and external processes may need to be reinforced (the single sentence pointing that out is read, forgotten, and then confusion later follows). May want to completely avoid QM at this point. See if can mention without raising further distracting questions.

@@

In this context, we define a process \emph{deterministic} if the state at a given time uniquely identifies states at future times, and \emph{reversible} if it identifies states at past times. This means that, in general, the evolution of the unstated part is non-deterministic and/or non-reversible in the sense that the state of the system is not enough to determine its evolution. For example, the overall motion of a fluid in a pipe can be considered deterministic even if the motion of each molecule is not. The motion is deterministic and/or reversible only as far as the state is concerned. With this in mind, we introduce the following:

@@

As we have hinted, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the trajectories of the parts are not (those fall within the unstated part). In fact: we \emph{cannot} assume trajectories and states are always defined at all for the unstated part. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal independent state and evolution to each part. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.

@@

>>>>

% Status: Conceptual work done. Text ready for review. Incorporated Dave feedback

@@

In this context, we define a process \emph{deterministic} if the state at a given time uniquely identifies states at future times, and \emph{reversible} if it identifies states at past times. While this is a common enough definition, we need to be clear how this applies to the unstated part. That is: the evolution of the unstated part is \emph{always} non-deterministic (and non-reversible) in the sense that the state of the system does not determine its evolution. For example, suppose we define the state as position and momentum of the center of mass of a cannonball. Suppose that the process is deterministic on that state. What does it tell us about the cannonball temperature, or about the motion of each of its atoms? Nothing. In this sense, the unstated part is non-deterministic and non-reversible. Could we extend the state and the laws of evolution to account for temperature? Yes, but that would be a different process defined on a different state. Does it mean the unstated part is always evolving chaotically? Not at all: the temperature may remain constant throughout the motion of the cannonball. Yet we wouldn't know, since we are not studying it: the motion is deterministic and/or reversible only as far as the state is concerned. With this in mind, we introduce the following:

@@

As we have hinted, sometimes the state is identified by a distribution (either statistical or actual). Even in this case, the state can be deterministic and reversible. That is: given the distribution at one time we can determine the distribution at future times. The shape and the parameters of the distribution can be deterministic, even if the trajectories of the parts are not (those fall within the unstated part). In fact: we \emph{cannot} assume trajectories and states are always defined at all for the unstated part, as this includes also the unknown unknowns. Will return to this aspect when discussing quantum systems.

@@

% Muon example. Consider a muon and its decay into an electron and two neutrinos: it is clear that the three outgoing particles have a state and trajectory of their own, it is clear that the resulting total mass and energy came from the muon. Yet, before the decay, we cannot ascribe an internal independent state and evolution to each part. The state of a muon is not some combination of the state of an electron and two neutrinos. That is: the unstated part is not just microstates.
